---
title: "Get Started with `bigsimr` Julia Version"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{tidyverse, patchwork}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

## Installation

Please see the [Installation and Setup](https://schisslergroup.github.io/bigsimr/articles/install-bigsimr.html) page for getting `bigsimr` up and running. `bigsimr` requires a working version of Python that can be found in the path variable. It is also required that the version of python has `numpy` and `jax` installed.

## Usage

```{r setup}
## devtools::install_github("SchisslerGroup/bigsimr", ref="julia")
library(bigsimr)
Sys.setenv(JULIA_NUM_THREADS=parallel::detectCores())
## Either let JuliaCall find your installation of Julia or set manually
bs <- bigsimr_setup(JULIA_HOME = "/usr/local/bin/")
dist <- distributions_setup() # Julia distributions
library(tidyverse)
library(patchwork)
```

We're going to show the basic use and syntax of `bigsimr` by using the New York air quality data set (`airquality`) included in the R `datasets` package. We will focus specifically on the temperature (degrees Fahrenheit) and ozone level (parts per billion).

```{r}
df <- airquality %>%
  select(Ozone, Temp) %>%
  drop_na()

glimpse(df)
```

Let's look at the joint distribution of the Ozone and Temperature

```{r}
p0 <- ggplot(df, aes(Temp, Ozone)) +
  geom_point(size = 1) +
  theme(legend.position = "none") +
  labs(x = "Temperature")

pTemp <- ggplot(df, aes(Temp)) + 
  geom_density() +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank())

pOzone <- ggplot(df, aes(Ozone)) + 
  geom_density() + 
  theme(axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank()) +
  coord_flip() 

pTemp + plot_spacer() + p0 + pOzone + 
  plot_layout(widths = c(3,1), heights = c(1, 3))
```

We can see that not all margins are normally distributed; the ozone level is highly skewed. Though we don't know the true distribution of ozone levels, we can go forward assuming that it is log-normally distributed.

To simulate observations from this joint distribution, we need to estimate the correlation and the marginal parameters.

### Estimating Correlation

To estimate the correlation, we can use `R`'s function `cor` or we can use the convenience function `bigsimr::cor_fast` which can estimate Pearson, Spearman, or Kendall correlation using the fastest methods available from other packages.

```{r}
p <- bs$cor(as.matrix(df), bs$Pearson)
p
```

### Defining Marginal Distributions

Next we can estimate the marginal parameters. Assuming that `Temperature` is normally distributed, it has parameters

```{r}
mu_temp <- mean(df$Temp)
sd_temp <- sd(df$Temp)
```

and assuming that `Ozone` is log-normally distributed, it has parameters

```{r}
mu_ozone <- mean(log(df$Ozone))
sd_ozone <- sqrt(mean((log(df$Ozone) - mean(log(df$Ozone)))^2))
```

Finally we take the parameters and put them into an `alist` of margins.

```{r}
margins <- c(
  dist$Normal(mu_temp, sd_temp),
  dist$LogNormal(mu_ozone, sd_ozone)
)
margins
```

```{r}
D <- bs$MvDistribution(p, margins, bs$Pearson)
D
```

Right now the marginal parameters must be filled in manually with the quantile function of the distribution. If the margins are homogeneous, then some automation can be implemented.

### Correlation Bounds

Given a list of margins, the theoretical lower and upper correlation coefficients can be estimated.

```{r}
bounds <- bs$cor_bounds(D)
bounds$lower
bounds$upper
```

```{r}
bounds <- bs$pearson_bounds(D)
bounds$lower
bounds$upper
```

Since our estimated correlation is within the theoretical bounds (assuming the distributions are correct), we should be able to achieve a similar correlation in the simulated data.

### Simulating Multivariate Distributions

Let's now simulate 10,000 observations from the joint distribution using `bigsimr::rvec`

```{r}
x <- bs$rvec(10000L, p, margins)
head(x)
```

and visualize the joint distribution

```{r}
p1 <- df_sim %>%
  ggplot(aes(Temp, Ozone)) +
  geom_density_2d_filled() +
  theme(legend.position = "none") +
  labs(x = "Temperature", y = "Ozone") +
  scale_y_continuous(limits = c(0, max(df$Ozone))) +
  scale_x_continuous(limits = range(df$Temp))

p1Temp <- ggplot(df_sim, aes(Temp)) + 
  geom_density(alpha = 0.5, fill = "lightseagreen") +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank())

p1Ozone <- ggplot(df_sim, aes(Ozone)) + 
  geom_density(alpha = 0.5, fill = "lightseagreen") + 
  theme(axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank()) +
  coord_flip() 

p1Temp + plot_spacer() + p1 + p1Ozone + 
  plot_layout(widths = c(3,1), heights = c(1, 3))
```

## Comparison to Uncorrelated Samples

We can compare the bivariate distribution above to one where no correlation is taken into account.

```{r}
df_sim2 <- data.frame(
  Ozone = rlnorm(10000, mle_mean(df$Ozone), mle_sd(df$Ozone)),
  Temp = rnorm(10000, mean(df$Temp), sd(df$Temp))
)
```

```{r, echo=FALSE, warning=FALSE}
p2 <- df_sim2 %>%
  ggplot(aes(Temp, Ozone)) +
  geom_density_2d_filled() +
  theme(legend.position = "none") +
  labs(x = "Temperature", y = "Ozone") +
  scale_y_continuous(limits = c(0, max(df$Ozone))) +
  scale_x_continuous(limits = range(df$Temp))

p2Temp <- ggplot(df_sim2, aes(Temp)) + 
  geom_density(alpha = 0.5, fill = "lightseagreen") +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank())

p2Ozone <- ggplot(df_sim2, aes(Ozone)) + 
  geom_density(alpha = 0.5, fill = "lightseagreen") + 
  theme(axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank()) +
  coord_flip() 

p2Temp + plot_spacer() + p2 + p2Ozone + 
  plot_layout(widths = c(3,1), heights = c(1, 3))
```

Notice that this uncorrelated bivariate distribution has the same marginal distributions as the correlated example above.

# Pearson matching

## Correlation conversion

```{r}
ps <- bs$cor(as.matrix(df), bs$Spearman)
ps
```

```{r}
bs$cor_bounds(margins[1], margins[2], bs$Spearman)
```

```{r}
D2 <- bs$MvDistribution(ps, margins, bs$Spearman)
x2 <- bs$rand(D2, 1000000L)
bs$cor(x2, bs$Spearman)
```

```{r}
pp <- bs$cor_convert(ps, bs$Spearman, bs$Pearson)
D3 <- bs$MvDistribution(pp, margins, bs$Pearson)
x3 <- bs$rand(D3, 1000000L)
bs$cor(x3, bs$Pearson)
bs$cor(x3, bs$Spearman)
```

```{r}
D4 <- bs$pearson_match(D2);
bs$cor(D4)
```

```{r}
x4 <- bs$rand(D4, 1000000L)
bs$cor(x4, bs$Pearson)
```

```{r}
bs$cor(bs$rand(D, 1000000L))
p
```

```{r}
D5 <- bs$pearson_match(D)
x5 <- bs$rand(D5, 1000000L)
bs$cor(x5)
```

```{r}
m <- as.matrix(brca)
str(m)
```

```{r}
fit_mom <- function(x) {
  m <- mean(x)
  s <- sd(x)
  r <- m^2 / (s^2 - m)
  p <- m / s^2
  dist$NegativeBinomial(r, p)
}
margins <- apply(m, 2, fit_mom)
margins <- do.call(c, unname(margins))
```

```{r}
tau <- bs$cor(m, bs$Spearman)
rho <- bs$cor_convert(tau, bs$Spearman, bs$Pearson)
matrixcalc::is.positive.definite(rho)
```

```{r}
bs$rvec(10L, rho, margins)
```

```{r}
rho_tilde <- bs$cor_nearPD(rho)
matrixcalc::is.positive.definite(rho_tilde)
```

```{r}
x <- bs$rvec(10L, rho_tilde, margins)
str(x)
```
