---
title: Bivariate Gamma Correlation Bounds Comparison
author: |-
    A. Grant Schissler\
    Department of Mathematics & Statistics\
    The University of Nevada, Reno. \
    aschissler@unr.edu
date: '`r Sys.Date()`'
output:
    html_document:
      toc: yes
      toc_depth: 2
bibliography: /Users/alfred/Dropbox/bib/library.bib
abstract:
    Quick study checking the `bigsimr` Pearson matching scheme while checking whether the methods obtain pairwise Frechet bounds.
---

```{r LoadLib, echo = TRUE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide"}
library(ggplot2)
library(tidyverse)
library(knitr)
library(dplyr)
knitr::opts_chunk$set(cache=TRUE, autodep=TRUE, cache.comments=TRUE)
set.seed(06032020)
## devtools::install_github( "adknudson/bigsimr", ref = 'develop')
library(bigsimr)
## devtools::install_github( "superdesolator/NORTARA" )
library(nortaRA)
## to parallize
library(parallel)
CORES <- parallel::detectCores() - 1
```

# Configure global parameters

```{r, globalParams, results = "hide"}
## bivariate gamma
d <- 2
shape1 <- 5
shape2 <- 5
margins <- list(
    list("gamma", shape = shape1),
    list("gamma", shape = shape2)
)

## for evaluation of obtaining the bounds
gridSteps <- 1e2
N <- 1e5

## storing the results
boundsDat <- NULL
simRes <- NULL

## parallelize outside of bigsimr
cl <- makeCluster(CORES)
clusterEvalQ(cl, library("bigsimr"))
```

```{r runSim}
runSim <- function( r, N, margins, cores, simMethod, corType, invcdfnames = NULL) {
    d <- length(margins)
    print( paste0( 'Working on true rho of ', r ) )
    rho <- matrix(r, nrow = d, ncol = d)
    diag(rho) <- 1.0
    if (simMethod == 'bigsimr') {
        simData  <- as.data.frame( rvec(n = N, rho = rho, params = margins, type = corType, cores = cores) )
    }
    if (simMethod == 'nortaRA') {
        simData <- nortaRA::gennortaRA(n = N, cor_matrix = rho, invcdfnames, margins)
    } 
    tmpSimData <- data.frame( shape1 = margins[[1]]$shape, shape2 = margins[[2]]$shape, trueRho = r, estRho = fastCor(simData, method = type)[ 1, 2], method = simMethod , type = corType)
    return(tmpSimData)
}
## runSim( r = 0.5, N = 1e2, margins = margins, cores = 1, simMethod = 'bigsimr', corType = 'pearson')
```

# Compute bounds under each type of dependency

## Pearson correlation bounds

```{r, boundsPearson}
type <- 'pearson'
## Using GSC method of Demirtas2011
(tmpBounds <- computeCorBounds( params = margins, cores = CORES, type = type, reps = N ))
tmpBoundsDat <- data.frame( shape1 = shape1, shape2 = shape2, type = type, upper = tmpBounds$upper[ 1, 2 ], lower = tmpBounds$lower[ 1, 2 ] )
boundsDat <- rbind( boundsDat, tmpBoundsDat )
```

## Spearman correlation bounds

```{r, boundsSpearman}
type <- 'spearman'
## Using GSC method of Demirtas2011
(tmpBounds <- computeCorBounds( params = margins, cores = CORES, type = type, reps = N ))
tmpBoundsDat <- data.frame( shape1 = shape1, shape2 = shape2, type = type, upper = tmpBounds$upper[ 1, 2 ], lower = tmpBounds$lower[ 1, 2 ] )
boundsDat <- rbind( boundsDat, tmpBoundsDat )
```

## Kendall correlation bounds

```{r, boundsKendall, eval=TRUE}
type <- 'kendall'
## Using GSC method of Demirtas2011
(tmpBounds <- computeCorBounds( params = margins, cores = CORES, type = type, reps = N ))
tmpBoundsDat <- data.frame( shape1 = shape1, shape2 = shape2, type = type, upper = tmpBounds$upper[ 1, 2 ], lower = tmpBounds$lower[ 1, 2 ] )
boundsDat <- rbind( boundsDat, tmpBoundsDat )
```

# Simulating and checking the empirical bounds obtainment

## `bigsimr` simulations

### Pearson

```{r bigsimrPearson, eval = TRUE, results = 'hide'}
simMethod <- 'bigsimr'
type <- 'pearson'

## now show that our simulation methods works
tmpLower <- boundsDat$lower[ boundsDat$type == type ]
tmpUpper <- boundsDat$upper[ boundsDat$type == type ]
## rVec <- c(-0.4, 0.1, 0.5)
## rVec <- seq( tmpLower, tmpUpper, length.out = gridSteps)
## loose on the boundary for now
epsilon <- 1e-2
rVec <- seq( tmpLower+epsilon, tmpUpper-epsilon, length.out = gridSteps ) 

clusterExport(cl, varlist = c( 'simMethod', 'type') )
tmpSimRes <- as.data.frame( do.call( "rbind", parSapply( cl = cl, rVec, FUN = runSim, N = N, margins = margins, cores = 1, simMethod = simMethod, corType = type, simplify = FALSE) ) )
simRes <- rbind( simRes, tmpSimRes )

```

### Spearman

```{r bigsimrSpearman, eval = TRUE, results = 'hide'}
type <- 'spearman'
simMethod <- 'bigsimr'

## now show that our simulation methods works
tmpLower <- boundsDat$lower[ boundsDat$type == type ]
tmpUpper <- boundsDat$upper[ boundsDat$type == type ]
## rVec <- c(-0.4, 0.1, 0.5)
## rVec <- seq( tmpLower, tmpUpper, length.out = gridSteps)
## loose on the boundary for now
epsilon <- 1e-2
rVec <- seq( tmpLower+epsilon, tmpUpper-epsilon, length.out = gridSteps ) 

clusterExport(cl, varlist = c( 'simMethod', 'type') )
tmpSimRes <- as.data.frame( do.call( "rbind", parSapply( cl = cl, rVec, FUN = runSim, N = N, margins = margins, cores = 1, simMethod = simMethod, corType = type, simplify = FALSE) ) )
simRes <- rbind( simRes, tmpSimRes )

```

### Kendall

```{r bigsimrKendall, eval = TRUE}
type <-'kendall'
simMethod <- 'bigsimr'

## now show that our simulation methods works
tmpLower <- boundsDat$lower[ boundsDat$type == type ]
tmpUpper <- boundsDat$upper[ boundsDat$type == type ]
## rVec <- c(-0.4, 0.1, 0.5)
## rVec <- seq( tmpLower, tmpUpper, length.out = gridSteps)
## loose on the boundary for now
epsilon <- 1e-2
rVec <- seq( tmpLower+epsilon, tmpUpper-epsilon, length.out = gridSteps ) 

clusterExport(cl, varlist = c( 'simMethod', 'type') )
tmpSimRes <- as.data.frame( do.call( "rbind", parSapply( cl = cl, rVec, FUN = runSim, N = N, margins = margins, cores = 1, simMethod = simMethod, corType = type, simplify = FALSE) ) )
simRes <- rbind( simRes, tmpSimRes )
```

## `nortaRA` simulations

```{r nortaRAsetup, eval = TRUE}
## slightly different input format than bigsimr
invcdfnames <- c("qgamma","qgamma")
paramslists <- list(
    m1 = list(shape = shape1 ),
    m2 = list(shape = shape2 )
)
```

### Pearson

```{r nortaRApearson, eval = TRUE, warning=FALSE, results='hide'}
type <- 'pearson'
simMethod <- 'nortaRA'

## now show that our simulation methods works
tmpLower <- boundsDat$lower[ boundsDat$type == type ]
tmpUpper <- boundsDat$upper[ boundsDat$type == type ]
## rVec <- c(-0.4, 0.1, 0.5)
## rVec <- seq( tmpLower, tmpUpper, length.out = gridSteps)
## loose on the boundary for now
epsilon <- 1e-2
rVec <- seq( tmpLower+epsilon, tmpUpper-epsilon, length.out = gridSteps ) 

## drop N for this one
tmpN <- 1e4

## runSim( r = 0.4, N = tmpN, margins = paramslists, cores = 1, simMethod = simMethod, corType = type, invcdfnames = invcdfnames)
clusterExport(cl, varlist = c( 'simMethod', 'type') )
tmpSimRes <- as.data.frame( do.call( "rbind", parSapply( cl = cl, rVec, FUN = runSim, N = N, margins = paramslists, cores = 1, simMethod = simMethod, corType = type, invcdfnames = invcdfnames, simplify = FALSE) ) )
simRes <- rbind( simRes, tmpSimRes )
```

# Check the performance around the bounds

```{r obtainment}
## numerically compare to the bounds

```

# Visualize the results

```{r empVspecPlot, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, fig.path='fig/plot-', out.width= '75%', fig.align='center', fig.ext='pdf', fig.cap = 'Plot of the specified versus empirical correlations for standard bivariate gamma data with shape1 = shape2 = 5 generated by various approaches. Empirical correlations are the estimated correlations from the simulated samples.'}
p0 <- ggplot(data = simRes, aes(x = trueRho, y = estRho, color = factor(method)) )
p0 + geom_point() + facet_grid(. ~ type ) + geom_abline(slope = 1, intercept = 0 )
```

```{r cleanUP, echo = FALSE}
## When you're done, clean up the cluster
## stopCluster(cl)
```
